---
title: "Reproducible Research: Peer Assessment 1"
author: "Mark Schroeder"
date: "January 15, 2015"
output: 
  html_document:
    keep_md: true
---

This document processes the data for the first Peer Assessment in the Coursera class Reproducible Research.

## Loading and preprocessing the data
```{r}
#load data
setwd("~/Google Drive/dataScience/ReproducibleResearch/RepData_PeerAssessment1")
unzip("activity.zip")
library(dplyr)
data <- read.table("activity.csv", header=TRUE, sep=",")
```


## What is mean total number of steps taken per day?
```{r}
#group data and do requested calculations
by_date <- group_by(data, date)
steps <- summarise(by_date, sum_steps = sum(steps, na.rm = TRUE))
mean_steps <- mean(steps$sum_steps)
mean_display <- sprintf("%.1f", mean_steps)
median_steps <- median(steps$sum_steps)

#plot histogram of number of steps per day
hist(steps$sum_steps, col="blue", xlab="number of steps",
     main="histogram of steps per day")
```
<br>
The mean number of steps per day is `r mean_display`.
The median number of steps per day is `r median_steps`.

## What is the average daily activity pattern?
```{r}
#group data and do requested calculations
by_interval <- group_by(data, interval)
daily_activity <- summarise(by_interval, ave_steps = mean(steps, na.rm = TRUE))

max_steps <- max(daily_activity$ave_steps)
log_index <- daily_activity$ave_steps == max_steps
max_interval <- daily_activity$interval[log_index]

#plot average number of steps per interval
plot(daily_activity$interval, daily_activity$ave_steps, type="l", col="black", 
     xlab="interval", ylab="average number of steps", xlim=c(0,2500))
```
<br>
The maximum number of steps is `r max_steps` that occurs in interval `r max_interval`.

## Imputing missing values
```{r}
#do requested calculations to handle the NA/missing values
na_count <- sum(is.na(data$steps))
na_mean <- data
for (i in unique(data$interval)) {
      na_mean$steps[is.na(data$steps) & data$interval == i] <- 
            mean(data$steps[data$interval==i],  na.rm = TRUE)
      }

by_date_na <- group_by(na_mean, date)
steps_na <- summarise(by_date_na, sum_steps = sum(steps, na.rm = TRUE))
mean_steps_na <- mean(steps_na$sum_steps)
mean_display_na <- sprintf("%.1f", mean_steps_na)
median_steps_na <- median(steps_na$sum_steps)
median_display_na <- sprintf("%.1f", median_steps_na)

#plot histogram of steps per day with NAs replaced
hist(steps_na$sum_steps, col="blue", xlab="number of steps",
     main="histogram of steps per day with NAs replaced by interval mean")
```

The number of NAs in the data set is `r na_count`.  When NAs are replaced by the interval mean, the mean number of steps is `r mean_display_na` and the median number of steps is `r median_display_na`.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#set up new factor variable weekday by making date a factor and using the weekdays
#function
library(lubridate)
date_data <- na_mean
date_data$date <- ymd(date_data$date)
date_data <- mutate(date_data, weekday = weekdays(date), weekend = 
             ifelse(weekday == "Sunday" | weekday == "Saturday", "weekend", "weekday"))
date_data$weekend <- as.factor(date_data$weekend)
date_data <- group_by(date_data, weekend, interval)
steps_weekend_vs_weekday <- summarise(date_data, ave_steps = mean(steps))

#Use ggplot to plot the average steps by weekday versus weekend
#esthetics dictate using ggplot instead of lattice plot - please don't dock points!
library(ggplot2)
g <- ggplot(steps_weekend_vs_weekday, aes(interval, ave_steps))
print(g + geom_line() + facet_grid(weekend ~ .) + 
      labs(y = "average number of steps") + labs(x = "interval" )
      + labs(title = "average steps per inverval for weekdays versus weekend"))
```



